Themis-ml:

### 2017-09-06 paper/main.tex
added paper and notebook evaluation
had to modify some of the metrics, and
also added a fairness-aware meta estimator
for convenience

\subsection{Experimental Procedure}

To assess the extent to which (i) a model trained on these data mirrors these PD
\emph{credit risk} distributions and (ii) fairness-aware techniques can reduce
these methods, we used \emph{mean difference} to measure model fairness and
\emph{AUC} to measure model utility. For this experiment we specify five
conditions:

\begin{itemize}
  \item Baseline (\emph{B}): Train a model on all available input variables in
        the German Credit dataset, including protected attributes.
  \item Remove Protected Attribute (\emph{RPA}): Train a model on input variables
        without protected attributes. This is the naive fairness-aware approach.
  \item Relabel Target Variable (\emph{RTV}): Train a model using the
        \emph{relabelling} fairness-aware method.
  \item Counterfactually Fair Model (\emph{CFM}): Train a model using the
        \emph{additive counterfactually fair} method.
  \item Reject-option Classification (\emph{ROC}): Train a model using the
        \emph{reject-option classification} method.
\end{itemize}

For each of these conditions, we train LogisticRegression, DecisionTree, and
RandomForest model types using 10-fold cross validation, generate train and test
predictions, and compute \emph{AUC} and \emph{mean difference} metrics for each
train-test pair. We then compute the mean these metrics for each condition and
model type.

\subsection{Measuring and Mitigating Potentially Discriminatory Predictions}

\begin{figure}[h]
\centering
\includegraphics[width=8.5cm]{IMG/fairness_aware_comparison.png}
\caption{
  \textbf{Comparison of Fairness-aware Methods} using LogisticRegression,
  DecisionTree, and RandomForest as base estimators as measured by AUC and
  mean difference evaluated on test set predictions.}
\label{figure-1-comparison-fa-methods}
\end{figure}

\textbf{Figure 1} suggests that in the case of LogisticRegression, the baseline
model \emph{B} does indeed mirror the PD patterns found in the true target
variable. Furthermore, each of the fairness-aware method appear to have
the desired effect of reducing \emph{mean difference}, but to variance degrees
depending on the method and protected attribute. In the \emph{female} context,
where there appear to be the least PD (mean difference of \(7.48\%\)), the
reductive effect of the fairness-aware methods does not appear to be as
large as in the \emph{foreign worker} and \emph{age below 25} contexts.

The lack of reduction in \emph{mean difference} between \emph{B} and \emph{RPA}
with respect to the \emph{foreign worker} protected attribute and
LogisticRegression model illustrates the observation made by
\cite{kusner2017counterfactual} that removing protected attributes from the
training data does not necessarly prevent the learning algorithm address from
mirroring PD patterns in the data.

However, as the sizeable reduction in \emph{mean difference} between \emph{B}
and \emph{RPA} with respect to the \emph{age below 25} protected attribute and
LogisticRegression model shows, removing protected attributes can make models
more fair while also retaining predictive power.

An interesting thing to note here is that the \emph{additive counterfactually
fair} method actually increases \emph{mean difference} for DecisionTrees and
RandomForests across all protected attribute contexts. A possible explanation
behind this observation is that certain assumptions made by the \emph{additive
counterfactually fair} method are not suitable for non-linear learning
algorithms, but this is an open question worth future inquiry.

\subsection{The Fairness-utility Tradeoff}

Just as the bias-variance tradeoff has become a useful diagnostic tool to guide
machine learning research and application, the fairness-utility tradeoff can
help machine learning practitioners and researchers understand how effective
particular fairness-aware methods might be in particular data contexts.

\begin{figure}[h]
\centering
\includegraphics[width=8.5cm]{IMG/fairness_utility_tradeoff.png}
\caption{
  \textbf{Correlation between AUC and Mean Difference} for each experimental
  condition, across all model types (LogisticRegression, DecisionTree,
  RandomForest). 95\% confidence intervals are provided for the pearson \(R\)
  correlation metric.
  }
\label{figure-2-fairness-utility-tradeoff}
\end{figure}

\textbf{Figure 2} demonstrates an initial attempt at visualizing the the
fairness-utility tradeoff, in this case as measured by \emph{mean difference}
and \emph{AUC} respectively. These results suggest that the relationship between
fairness and utility is noisy, however there does seem to be a positive
correlation between \emph{mean difference} and \emph{AUC}, or a negative
correlation between fairness and utility (since lower scores are better
for \emph{mean difference} and higher scores are better for \emph{AUC}).

Depending on one's use cases, analyses like this might be prove to be a useful
guide for figuring out what kinds of methods are robust in the sense that
one can reduce PD predictions with little to no adverse impact on predictive
performance.

@@

In this paper, we describe and evaluate an FMLI in the classification context
where we consider only a single binary protected class variable and a binary
target variable. More work needs to be done to generalize FMLIs to the multi-
classification, regression, and multiple protected classes contexts.

Furthermore, many basic questions about model tuning, model evaluation, and
model selection in relation to the fairness-utility tradeoff remain open. For
instance, what might be some reasonable ways to aggregate utility and
performance metrics in order to find the optimal set of hyperparameters?
Additionally, little is understood about the question of the composability of
fairness-aware methods. In other words, when different techniques are used
together in sequence, are the resulting discrimination reductions additive or
otherwise?

>>>>

\subsection{Measure and Mitigate Potentially Discriminatory Predictions}
To assess the extent to which a model trained on these data mirrors these PD
\emph{credit risk} distributions, we use the following experimental procedure:

\begin{enumerate}
  \item \textbf{Models:} LogisticRegression, DecisionTree, RandomForest
  \item train model using leave-one-out 10-fold cross validation
  \item measure PD predictions in a machine learning algorithm's predictions.
  \item reduce PD predictions using fairness-aware techniques.
  \item diagnose the fairness-utility tradeoff in a particular data context.
\end{enumerate}
train a logistic regression model using
leave-one-out 10-fold cross validation, generate predictions on the training and
test folds, and compute \emph{AUC} and \emph{mean difference} for each
train-test partition. We then compute the mean of these performance metrics
over a 1-dimensional grid of L2 regularization hyperparemeter settings.

\begin{figure}[h]
\centering
\includegraphics[width=8.5cm]{IMG/fairness_aware_comparison.png}
\caption{
  \textbf{Comparison of Fairness-aware Methods}
  % are shown for each of the three protected class contexts: \emph{female},
  % \emph{foreign worker}, and \emph{age below 25}. Left panels show training and
  % test \emph{AUC} scores (higher is better) of an L2-regularized Logistic
  % Regression model. The \(C\) parameter is the inverse regularization strength,
  % meaning that smaller values increase the regularization penalty. Model
  % complexity increases as \(C\) increases, leading to the overfitting regime at
  % \(C > 10^{-1}\). Right panels show training and test \emph{mean difference}
  % scores (lower is better).
  }
\label{figure-1-comparison-fa-methods}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=8.5cm]{IMG/fairness_utility_tradeoff.png}
\caption{
  \textbf{Diagnosing the Fairness-utility Tradeoff}
  % are shown for each of the three protected class contexts: \emph{female},
  % \emph{foreign worker}, and \emph{age below 25}. Left panels show training and
  % test \emph{AUC} scores (higher is better) of an L2-regularized Logistic
  % Regression model. The \(C\) parameter is the inverse regularization strength,
  % meaning that smaller values increase the regularization penalty. Model
  % complexity increases as \(C\) increases, leading to the overfitting regime at
  % \(C > 10^{-1}\). Right panels show training and test \emph{mean difference}
  % scores (lower is better).
  }
\label{figure-2-fairness-utility-tradeoff}
\end{figure}

% \begin{figure}
%   \centering
%   \begin{subfigure}[t]{1in}
%     \centering
%     \includegraphics[width=6cm]{IMG/logistic_regression_validation_curve.png}
%     \caption{Arabic numerals}\label{fig:1a}
%   \end{subfigure}
%   \quad
%   \begin{subfigure}[t]{1in}
%     \centering
%     \includegraphics[width=3cm]{IMG/logistic_regression_fairness_utility_tradeoff.png}
%     \caption{Arabic numerals}\label{fig:1b}
%   \end{subfigure}
%   \caption{Capital Roman numerals}\label{fig:1}
% \end{figure}

In the machine learning field we often talk about the bias-variance tradeoff,
which refers to the observation that simple models exhibit high bias and low
variance, i.e. they ``underfit'' the data by consistently performing poorly on
both the test and training set across many instantiations of the model.

However, as model complexity increases (for example, by introducing first order
polynomial terms into space), models begin to exhibit low bias and high
variance. This means that they ``overfit'' the data by performing well on the
training set but poorly on the test set.

In practice, we often seek to find the optimal model specification that
maximizes/minimizes test set performance/error but also minimizes the gap
between test set and training set performance/error.

A concept that is related but distinct from the bias-variance tradeoff is the
fairness-utility tradeoff, which is the observation that fair models tend to
make less accurate predictions \cite{kamiran2012data, dwork2012fairness}. Just
as the bias-variance tradeoff has become a useful diagnostic tool to guide
machine learning research and application, the fairness-utility tradeoff can
help .

@@

In this paper, we describe FMLIs in the classification context where we consider
only a single binary protected class variable. More work needs to be done to
generalize FMLIs to the multi-classification, regression, and multiple protected
classes contexts. Additionally, little is understood about the question of the
composability of fairness-aware methods. In other words, when different
techniques are used together in sequence, are the resulting discrimination
reductions additive or otherwise?
